SunPy - Python for Solar Physics

### 2014-01-27 2-1-Map.tex

\subsection{Map}
Map is a 2D spatial data type, primarily used for images of the Sun and 
inner heliosphere. It provides a wrapper around a data array (\texttt{numpy} 
\texttt{ndarray}) and gives easy access to standard meta data in the header of 
the image, via the \texttt{.meta} attribute as well as some convenience 
properties for more standard meta data, \textit{i.e.} \texttt{.instrument}, 
\texttt{.wavelength} or \texttt{.coordinate\_system}. Demonstrated in Listing 
\ref{code:aia_1} is this meta data parsing and `pretty printing' of the Map.
The \texttt{Map} object provides convenience methods for many functions 
such as, rotation and re-sampling as well as convenience visualisation 
functions, providing an easy to use as well as powerful interface.
Listing \ref{code:aia_1} also demonstrates the quick plot functionality of 
\texttt{Map}, see Section \ref{subsec:Viz} for more discussion of plotting.

The design of the map submodule is such that each instrument or 
detector can subclass the parent \texttt{GenericMap} class to implement 
special meta data handling or other data specific functions. Each subclass 
of \texttt{GenericMap} can register with the \texttt{Map} factory class and 
by implementing a method that returns \texttt{True} if the meta data 
matches meta data for that instrument or detector, the \texttt{Map} factory 
will automatically return an instance of the specific \texttt{GenericMap} 
subclass. As of version 0.4 SunPy has \texttt{Map} specialisations for the 
following instruments: 
\textit{YOHKOH/SXT},\textit{SOHO/EIT} and \textit{LASCO}, \textit{RHESSI}, 
\textit{STEREO/EUVI} and \textit{COR}, \textit{HINODE/XRT},
\textit{PROBA2/SWAP}, \textit{SDO/AIA} and \textit{HMI}, 
and \textit{IRIS} SJI frames.

\texttt{GenericMap} is a subclass of \texttt{astropy.nddata.NDData} which is a 
generic wrapper around a \texttt{ndarray} with a \texttt{.meta} attribute, like 
\texttt{Map}. Having this inheritance structure with \texttt{astropy} provides 
potential for simple integration with other parts of \texttt{astropy} and other 
packages based on \texttt{astropy}, as \texttt{GenericMap} should form a 
drop-in replacement for \texttt{NDData}.

\begin{listing}[H]
\begin{minted}[bgcolor=bg]{pycon}

>>> import sunpy.map

>>> aiamap = sunpy.map.Map('aia_file.fits')
>>> aiamap
SunPy AIAMap
---------
Observatory: SDO
Instrument:  AIA
Detector:    AIA
Measurement: 171
Obs Date:    2011-03-19T10:54:00.34
dt:          1.999601
Dimension:   [1024, 1024]
[dx, dy] =   [2.400000, 2.400000]

array([[ 0.3125, -0.0625, -0.125 , ...,  0.625 , -0.625 ,  0.    ],
       [ 1.    ,  0.1875, -0.8125, ...,  0.625 , -0.625 ,  0.    ],
       [-1.1875,  0.375 , -0.5   , ..., -0.125 , -0.625 , -1.1875],
       ..., 
       [-0.625 ,  0.0625, -0.3125, ...,  0.125 ,  0.125 ,  0.125 ],
       [ 0.5625,  0.0625,  0.5625, ..., -0.0625, -0.0625,  0.    ],
       [ 0.5   , -0.125 ,  0.4375, ...,  0.6875,  0.6875,  0.6875]])

>>> aiamap.peek()
\end{minted}
\includegraphics[width=0.8\columnwidth]{aia_map_example}
\caption{Demonstration of the \texttt{AIAMap} specialisation of 
\texttt{GenericMap}. The Map is created from an \textit{AIA} FITS file and the 
key meta data and array overview is printed. Then a quick view plot is created 
by using the \texttt{.peek()} method.}
\label{code:aia_1}
\end{listing}

As well as providing the base classes the map submodule provides two 
collection classes, \texttt{CompositeMap} and \texttt{MapCube}, for 
temporally and spatially aligned data respectively.
\texttt{CompositeMap} provides methods for overlaying spatially aligned 
data, with support for visualisation of images and contour lines overlaid 
upon each other.
\texttt{MapCube} 
provides methods for animation of its series of \texttt{Map} objects. 
Listings~\ref{code:compmap_1} and \ref{code:mapcube_1} shows how to interact 
with these classes.
In addition, Listing~\ref{code:mapcube_2} demonstrates how the \texttt{MapCube} 
classes interacts with the \texttt{animation} module to save a video.

>>>>

\subsection{Map}
The purpose of the Map datatype is to represent 2D spatial data such as images of the Sun and 
inner heliosphere. It provides a wrapper around a data array and its associated
spatial coordinates and as well as meta data. The \texttt{Map} object provides 
convenience methods for many functions 
such as rotation and re-sampling as well as convenience visualisation 
functions, providing an easy to use as well as powerful interface. The \texttt{Map} object
also provides a convenient interface for loadings data from a variety of sources, typically
a FITS file as shown in Figure~\ref{code:aia_1}.

The architecture of the \texttt{Map} object consists of a template map called
\texttt{GenericMap} which is a subclass of \texttt{astropy.nddata.NDData}. \texttt{NDData} is a 
generic wrapper around a \texttt{ndarray} with a \texttt{.meta} attribute to store meta data.
As \texttt{NDData} is currently still in development, \texttt{GenericMap} does not make full
use of its capabilities but this inheritance structure provides 
for future integration of \texttt{astropy}. In order to provide instrument/
detector specific integration \textt{GenericMap} is subclassed on creation. 
Each subclass of \texttt{GenericMap} can register with the \texttt{Map} creation function (i.e. factory)
by implementing a method that returns \texttt{True} if matching meta data 
for that instrument or detector is found in the file header. The \texttt{Map} factory 
will then automatically return an instance of the specific \texttt{GenericMap} 
subclass. As of this version Sunpy has \texttt{Map} specialisations for the 
following instruments: 
\textit{YOHKOH/SXT},\textit{SOHO/EIT} and \textit{LASCO}, \textit{RHESSI}, 
\textit{STEREO/EUVI} and \textit{COR}, \textit{HINODE/XRT},
\textit{PROBA2/SWAP}, \textit{SDO/AIA} and \textit{HMI}, 
and \textit{IRIS} SJI frames.

The \texttt{Map} object stores standard meta data retrieved in the header of 
the image file, via the \texttt{.meta} attribute and provides convenience 
properties for more standard meta data, \textit{i.e.} \texttt{.instrument}, 
\texttt{.wavelength} or \texttt{.coordinate\_system}. 
Listing \ref{code:aia_1} also demonstrates the quick plot functionality of 
\texttt{Map}, see Section \ref{subsec:Viz} for more discussion of plotting.

\begin{listing}[H]
\begin{minted}[bgcolor=bg]{pycon}
>>> import sunpy.map
>>> aiamap = sunpy.map.Map('aia_file.fits')
>>> aiamap.peek()
\end{minted}
\includegraphics[width=0.8\columnwidth]{aia_map_example}
\caption{Demonstration of the \texttt{AIAMap} specialisation of 
\texttt{GenericMap}. The Map is created from an \textit{AIA} FITS file and the 
key meta data and array overview is printed. Then a quick view plot is created 
by using the \texttt{.peek()} method.}
\label{code:aia_1}
\end{listing}

As well as providing the base classes the map sub-package provides two 
collection classes, \texttt{CompositeMap} and \texttt{MapCube}, for 
temporally and spatially aligned data respectively.
\texttt{CompositeMap} provides methods for overlaying spatially aligned 
data, with support for visualisation of images and contour lines overlaid 
upon each other.
\texttt{MapCube} 
provides methods for animation of its series of \texttt{Map} objects. 
Figure~\ref{code:compmap_1} and \ref{code:mapcube_1} shows how to interact 
with these classes.
In addition, Figure~\ref{code:mapcube_2} demonstrates how the \texttt{MapCube} 
classes interacts with the \texttt{animation} module to save a video.

### 2014-01-27 2-2-Lightcurve.tex
subsection label corrected

Time series data and analyses are a fundamental aspect of solar physics 
for which many data sources are available. 
In recognition of this fact, SunPy provides a \texttt{LightCurve} object 
designed to provide a convenient and consistent interface for handling 
solar time series data. 
The main engine behind the \texttt{LightCurve} object is the \texttt{pandas} 
data analysis library. 
This library contains a large amount of functionality for manipulating 
and analysing time series data, making it an ideal basis for 
\texttt{LightCurve}.  
Like all other SunPy objects, \texttt{LightCurve} is a wrapper around 
a data object, which in this case is the \texttt{pandas} object.

Currently, the \texttt{LightCurve} object is compatible with the following 
data sources: the GOES X-ray Sensor (XRS), PROBA2/LYRA, 
and the SDO EUV Variability Experiment (EVE). 
For each of these instruments, a sub-class of the \texttt{LightCurve} object 
is initialised (e.g. \texttt{GOESLightCurve, LYRALightCurve}) which inherits 
from \texttt{LightCurve} but allows instrument-specific functionality to be 
included. 
Future developments will introduce support for additional instruments and data 
products. 
Since time series data is generally relatively small and there is no 
established standard as to how it should be stored and distributed, 
each SunPy \texttt{LightCurve} object provides the ability to download 
its own data in its constructor.

@@

A \texttt{LightCurve} object may be created using a range of different methods. 
Firstly, a \texttt{LightCurve} may be created for a specific instrument based 
on an input time range. In Listing \ref{code:goes_lc_code}, 
the LightCurve constructor searches remote sites for the GOES X-ray data 
specified by the time interval, downloads the required files and subsequently 
creates the object.

\begin{listing}[H]
\begin{minted}[bgcolor=bg]{pycon}
>>> from sunpy import lightcurve
>>> goes=lightcurve.GOESLightCurve.create('2011-06-07 06:00',
...                                       '2011-06-07 08:00')
>>> goes.peek()
\end{minted}
\includegraphics[width=14cm]{goes_lightcurve.pdf}
\caption{Creating a GOES lightcurve for the time interval 06:00 - 08:00 UT on 
2011 June 7 using a time range, and the result of the \texttt{peek()} method.}
\label{code:goes_lc}
\end{listing}

Alternatively, if the data file already exists on the local system, the 
\texttt{Lightcurve} object may instead be initialised using that file as input 
(see Listing \ref{code:lyra_lc}).

\begin{listing}[H]
\begin{minted}[bgcolor=bg]{pycon}
>>> from sunpy import lightcurve
>>> file='lyra_20110607-000000_lev2_std.fits'
>>> lyra=lightcurve.LYRALightCurve.create(file)
>>> lyra.peek()
\end{minted}
\includegraphics[width=14cm]{lyra_lightcurve.pdf}
\caption{Creating a LYRA lightcurve for the full day of 2011 June 7 using an 
existing data file, and the result of the \texttt{peek()} method}
\label{code:lyra_lc}
\end{listing}

Once the \texttt{Lightcurve} has been created it may be manipulated in 
a variety of ways in order to perform time series analysis. 
Continuing from Listing~\ref{code:lyra_lc}, we show a few simple 
examples of using a LYRA lightcurve.

In Listing~\ref{code:lyra_lc} we created a LYRA Lightcurve spanning the 
full day of 2011 June 7. 
To extract the sub-interval 06:00 - 07:00 UT into a new \texttt{Lightcurve} 
called \texttt{lyra\_zoom} we simply do as shown on 
Listing~\ref{code:lyra_truncate}.

\begin{listing}[H]
\begin{minted}[bgcolor=bg]{pycon}
>>> from sunpy.time import TimeRange
>>> trange=TimeRange('2011-06-07 06:00','2011-06-07 07:00')
>>> lyra_zoom = lyra.truncate(trange)
\end{minted}
\caption{Extracting a sub-interval from a Lightcurve.}
\label{code:lyra_truncate}
\end{listing}

Also from the example shown in Listing~\ref{code:lyra_lc}, direct access to 
the data and time information in the object is available through 
\texttt{lyra.data} and \texttt{lyra.data.index} respectively 
(see Listing~\ref{code:lyra_pandas}).

\begin{listing}[H]
\begin{minted}[bgcolor=bg]{pycon}
>>> print lyra.data
<class pandas.core.frame.DataFrame>
DatetimeIndex: 1496629 entries, 2011-01-28 00:00:00.110000 to 
2011-01-28 23:59:59.983993
Data columns (total 4 columns):
CHANNEL1    1496629  non-null values
CHANNEL2    1496629  non-null values
CHANNEL3    1496629  non-null values
CHANNEL4    1496629  non-null values
dtypes: float64(4)

>>> print lyra.data.index
<class pandas.tseries.index.DatetimeIndex>
[2011-01-28 00:00:00.110000, ..., 2011-01-28 23:59:59.983993]
Length: 1496629, Freq: None, Timezone: None
\end{minted}
\caption{Accessing the data and time axis in a Lightcurve}
\label{code:lyra_pandas}
\end{listing}

>>>>

Time series data and analyses are a fundamental aspect of solar physics 
for which many data sources are available. 
Sunpy therefore provides a \texttt{LightCurve} object 
designed to provide a convenient and consistent interface for handling 
time series data. 
The main engine behind the \texttt{LightCurve} object is the \texttt{pandas}\footnote{\url{http://pandas.pydata.org}} 
data analysis library. 
This library contains a large amount of functionality for manipulating 
and analysing time series data, making it an ideal basis for 
\texttt{LightCurve}.  
Like all other Sunpy objects, \texttt{LightCurve} is a wrapper around 
a data object, which in this case is the \texttt{pandas} object.

Currently, the \texttt{LightCurve} object is compatible with the following 
data sources: the GOES X-ray Sensor (XRS), PROBA2/LYRA, 
and the SDO EUV Variability Experiment (EVE). 
For each of these instruments, a sub-class of the \texttt{LightCurve} object 
is initialised (e.g. \texttt{GOESLightCurve, LYRALightCurve}) which inherits 
from \texttt{LightCurve} but allows instrument-specific functionality to be 
included. 
Future developments will introduce support for additional instruments and data 
products. 
Since time series data is generally relatively small and there is no 
established standard as to how it should be stored and distributed, 
each Sunpy \texttt{LightCurve} object provides the ability to download 
its own data in its constructor.

@@

A \texttt{LightCurve} object may be created using a range of different methods. 
Firstly, a \texttt{LightCurve} may be created for a specific instrument based 
on an input time range. In Figure~\ref{code:goes_lc_code}, 
the LightCurve constructor searches a remote site for the GOES X-ray data 
specified by the time interval, downloads the required files and subsequently 
creates the object.

\begin{Figure}[H]
\begin{minted}[bgcolor=bg]{pycon}
>>> from sunpy import lightcurve
>>> from sunpy.time import TimeRange
>>> goes=lightcurve.GOESLightCurve.create('2011-06-07 06:00',
...                                       '2011-06-07 10:00')
>>> goes.peek()
>>> trange=TimeRange('2011-06-07 06:00','2011-06-07 07:00')
>>> goes_zoom = goes.truncate(trange)
% the above does not work!
>>> resampled_goes = goes.data.resample("1H")
\end{minted}
\includegraphics[width=14cm]{goes_lightcurve.pdf}
\caption{Example usage of a GOES lightcurve for the time interval 06:00 - 08:00 UT on 
2011 June 7 using a time range, and the result of the \texttt{peek()} method.
A sub-interval is then extracted from the Lightcurve and finally the data is resampled
to 1 hour cadence.}
\label{code:goes_lc}
\end{Figure}

Alternatively, if the data file already exists on the local system, the 
\texttt{Lightcurve} object may instead be initialised using that file as input.
Once the \texttt{Lightcurve} has been created it may be manipulated in 
a variety of ways in order to perform time series analysis. 
Continuing from Figure~\ref{code:lyra_lc}, we show a few simple 
examples of using a LYRA lightcurve.

### 2014-01-27 5-Dev.tex

\section{Development and Community}
SunPy is a community developed library, it is designed and developed for and by 
the solar physics community. Not only is all the source code publicly available 
online under the permissive 2-clause BSD licence 
\footnote{\url{http://opensource.org/licenses/BSD-2-Clause}}, the whole 
development 
process is also online and open for anyone to contribute to.
The primary media by which SunPy's development is facilitated is the online 
service GitHub \footnote{\url{http://github.com}} which makes use of the 
distributed 
version control software \href{http://git-scm.com/}{git}.

The development workflow that GitHub enables is simple and transparent. The 
primary source code repository \url{http://github.com/sunpy/sunpy} can be 
`forked' by anyone with a GitHub account who can then modify their fork as they 
wish. Upon completion of any modifications it is possible for a user to create 
a 'Pull Request' (\url{http://github.com/sunpy/sunpy/pulls}) to the main SunPy 
repository where it can be reviewed by anyone and then accepted by one of the 
development team who have commit rights to the main SunPy repository.

Pull requests are a very powerful tool to maintain code quality, as each line 
of code that has been modified is highlighted for review. For code to be 
accepted into the SunPy repository three primary things have to have been 
completed:

\begin{enumerate}
	\item  The code has to follow 	
	\href{http://www.python.org/dev/peps/pep-0008/}{PEP8} the Python style 
	guidelines, this keeps all the code in SunPy (and 99\% of all other Python 
	code) styled in a consistent manner and easy to read.

	\item The code must be documented. SunPy's documentation is generated by 
	the Sphinx package, which is automatically run and hosted by 
	\url{http://readthedocs.org}.

	\item The code must contain full unit test coverage. SunPy code is well 
	tested and coverage is increasing all the time. All SunPy's tests are 
	automatically run for each `Pull Request' and branch by the 
	\href{http://travis-ci.org/sunpy/sunpy}{Travis-CI }	service, and a test 
	coverage record is kept by 
	\href{http://coveralls.io/r/sunpy/sunpy}{Coveralls}
\end{enumerate}

This kind of development model is widely used within the scientific Python 
community as well as buy a wide variety of other projects both open and closed 
source.

As well as having a open source and open development model SunPy has a active 
community, with lots of both on-topic and off-topic chat on the \#SunPy IRC 
channel on freenode.net and on the mailing lists. The project always welcomes 
new contributions of any scale.

Many of the larger features added to SunPy during it's life have been added 
with the aid of both ESA's Summer Of Code In Space (SOCIS) program and Google's 
Summer of Code (GSOC) program. Initial development of SunPy's VSO and HEK 
implementations were funded by the SOCIS program as well as a prototyped GUI 
and N-Dimensional datatype implementations. 2013 was the first year that SunPy 
obtained GSOC funding, through the Python Software Foundation (PSF). The two 
projects that were undertaken was the development of a HELIO package and a 
package for integration between the HEK and VSO, allowing the download of data 
from VSO relating to a HEK query. As well as the Database module, which 
provides a single query-able interface between local data and the VSO.

>>>>

\section{Development and Community}
SunPy is a community-developed library, designed and developed for and by 
the solar physics community. Not only is all the source code publicly available 
online under the permissive 2-clause BSD licence 
\footnote{\url{http://opensource.org/licenses/BSD-2-Clause}}, the whole 
development 
process is also online and open for anyone to contribute to.
SunPy's development makes use of the online 
service GitHub \footnote{\url{http://github.com}} and \href{http://git-scm.com/}{git} 
as its distributed version control software.

The continued success of an open source projects depends on many factors of which 
three of the most important are (1) utility and quality of the code, (2) documentation, and (3) an
active community \cite{bangerth2013}. Several tools (some specific to Python) are used by
Sunpy to make achieving these goals more accessible. To maintain high quality code, a 
transparent and collaborative development workflow made possible by Github is used (described
further below). Sunpy includes a number of unit tests which test the functionality 
of the code base. Additionally, Sunpy makes use of `continuous integration' provided by
Travis\footnote{\url{http://travis-ci.org}}, a process by which the addition of any new code 
automatically triggers a comprehensive review of all unit tests. If any single test
fails then the community is alerted before the code is accepted. The unit test coverage is
also monitored by the \href{http://coveralls.io}{Coveralls} service.

Quality documentation is
one of the most important factors determining the success of any software project. 
Thankfully powerful tools already exist in Python to support this thanks to native
Python's focus on its own documentation. Sunpy makes use of the Sphinx\footnote{\url{http://sphinx-doc.org}}
documentation generator. Sphinx uses reStructuredText as its markup language, 
an easy-to-read, what-you-see-is-what-you-get plaintext markup syntax. It supports
many output formats most notably HTML as well as PDF, and ePub and provides a rich
hierarchically structured view of in-code documentation strings. The Sunpy documentation 
s built automatically and hosted on Readthedocs \footnote{\url{http://readthedocs.org}} 
at \url{http://docs.sunpy.org}.

In order to maintain an active community communication is key.  The Sunpy community makes
use of a number of channels to communicate. For immediate communications, an active IRC chat
room (\#SunPy) is hosted on freenode.net. For more involved or less immediate needs such as
developer comments or discussion, a open mailing list is used hosted on Google groups. 
Discussion of bugs and bug tracking are provided directly on Github as well as new code
review discussions.

Many of the larger features in SunPy have been developed 
with the support of both the ESA's Summer Of Code In Space (SOCIS) and the Google
Summer of Code (GSOC). Initial development of SunPy's VSO and HEK 
implementations were funded by the SOCIS program as well as a prototyped GUI 
and N-Dimensional datatype implementations. Sunpy was supported by GSOC in 2013 
through the Python Software Foundation (PSF). It supported development of the HELIO and 
database sub-packages.

\subsection{Development Workflow}
The development workflow that GitHub enables is simple and transparent. The 
primary source code repository \url{http://github.com/sunpy/sunpy} can be 
`forked' by anyone with a GitHub account who can then modify their fork as they 
wish. Upon completion of any modifications it is possible for a user to create 
a 'Pull Request' (\url{http://github.com/sunpy/sunpy/pulls}) to the main SunPy 
repository in order to incorporate their modification into the Sunpy code base. 
Pull requests are reviewed by the Sunpy community before being accepted. Community consensus
is required to add new features.

Pull requests are a very powerful tool to maintain code quality, as each line 
of code that has been modified is highlighted for review and can be commented on. 
In order to maintain high quality code the following conditions must typically be met 
before code is accepted.

\begin{enumerate}
	\item  The code must follow 	
	\href{http://www.python.org/dev/peps/pep-0008/}{PEP8} the Python style 
	guidelines, this keeps all the code in Sunpy styled in a consistent manner and easy to read.

	\item All new features require documentation in the form of doc strings as well as user
	guides.

	\item The code must contain significant unit test coverage.
\end{enumerate}

This kind of development model is widely used within the scientific Python 
community as well as by a wide variety of other projects both open and closed 
source.
